Nature or Nurture? Collisionless Evolution of Galactic Disc-Halo Systems

### 2019-07-16 Thesis/Conclusion.tex
Semi-final presubmission draft

\chapter{Summary and Conclusions}\label{ch:Conclusion}

@@

\section{Interactions Between Stellar Disks and Dark Matter Halos}

@@

\section{Bar Formation in Cosmological Stellar Disks}
\subsection{Summary}

@@

\section{Vetical Structure in Cosmological Stellar Disks}
\subsection{Summary}

@@

>>>>

\chapter{Summary and Conclusions}\label{ch:conclusion}
\newpage

@@

\section{Disk-Halo Interactions}

@@

\section{Bar Formation}
\subsection{Summary}

@@

\section{Vertical Structure}
\subsection{Summary}

@@

This thesis considered a very specific approach to understanding a very specific concept: the collisionless evolution of systems with stellar disks and dark matter halos. It is tremendously humbling to note that we never even considered gas dynamics, a cosmology other than $\Lambda$CDM, or even systems with a strong, pre-existing bulge. In the end, we were able to make some elucidating comments on the nature of stellar dynamics in $\Lambda$CDM cosmology, but no more.

As simulation resolutions and computing power improve, we may be forced to revisit the fundamentals at the beginning of Chapter~\ref{ch:background}. Simulation codes may need to be redesigned to accommodate ever-increasing needs to relate them to observations. For this kind of work to be truly impactful, we need to mind all of its benefits and shortcomings amd apply it accordingly.

Although our work raises more questions than it answers, it represents a tangible leap forward, however small, in the field of Galactic dynamics.  We hope that our work continues to inspire new developments, new comparisons, and more questions. In particular, we hope to see Galactoseismology mature, and possibly tell us about the nature of the dark Universe in which we live. This is an incredibly exciting time for the field, one where we are proud to have played a small role.

### 2019-07-16 Thesis/Implementation.tex
Semi-final presubmission draft

\chapter{Implementing Disc Insertion Schemes in \textsc{Gadget-3}}\label{ch:implementation}

@@

The disk position, velocity, angular position, and rate of Euler angle change are recorded in \texttt{disk\_vars.txt}. The differences in these quantities from the last timstep are recorded in \texttt{disk\_dvars.txt}. A variety of other quantities are written out into other log files that are pretty self explanatory. Appply \texttt{grep} to the code if you are unsure what a particular log file is. There should be comments.

@@

The actual integration scheme is handled in \texttt{rigiddisk.c}. The functions are self-explanatory and implement routines for generating the Euler rotataion matrix. At runtime, the particles in the rigid disk have their relevant quantities (acceleration, torque, etc.) rotated into the frame where the disk axis is the $z$-axis. We solve the angular part of the disk dynamics in this frame using the implicit Euler method. For the actual timestep used in the integration method, make sure you use the time calculated in \texttt{update\_halo\_positions.c}. This is the actual time elapsed as formulated\footnote{You can check that you're using the right timestep by ensuring the total time elapsed corresponds to the time elapsed in that cosmology.}, not the drift or kick steps given for the conjugate positions and velocities. We apologize for the horrible mess that this section of code has become. Perhaps it would be instructive to write your own halo tracker.

@@

As with most code developed in academia, individual researchers are largely responsible for the development, application, and maintenance of their codes. Balancing these priorties often means sacrificing on certain optimizations. Here, we list some potential improvements to the existing algorithm that might be explored.

@@

In the paper, we use a fixed timestep for the rigid disk that is based on the center-of-mass acceleration timescale. Often, this timescale is too long for the implicit Euler scheme that we use to integrate the rigid body equations. This is why we introduced a fixed \texttt{TIMESTEP\_REDUCTION\_FACTOR}. In reality, the timestep should be determined from the current magnitude of the torque and angular velocity. We would recommnend experimenting with an adaptive timestep. One form that might be tried is,
\begin{equation}
\Delta t = \frac{2 \pi \alpha}{\vert \boldsymbol \omega \vert}
\end{equation}
where $\alpha$ is a constant for which lower values yield more accurate integation, and $\boldsymbol \omega$ is the angular velocity in the disk's body frame. There were a number of halos where we found the fixed timestep to be inadequate for capturing the angular evolution of the disk. This is especially an issue when the disk begins to settle in the host halo and sheds some of its angular velocity. An adaptive timestep would give us the low error required to resolve these scenarios as well as giving us efficiency when angular velocities are low.

We are also able to improve on the integration schemne. We use, Implicit Euler, a first order integration scheme. By modifying the code to make use of the fact that \textsc{Gadget-3} is already using a higher order leapfrog scheme, we may be able to improve the integration accuracy.

>>>>

\chapter{Implementing Disc Insertion Schemes in \textsc{Gadget-3}}\label{ch:implementation}
\newpage

@@

The disk position, velocity, angular position, and rate of Euler angle change are recorded in \texttt{disk\_vars.txt}. The differences in these quantities from the last timestep are recorded in \texttt{disk\_dvars.txt}. A variety of other quantities are written out into other log files that are pretty self explanatory. Apply \texttt{grep} to the code if you are unsure what a particular log file is. There should be comments.

@@

The actual integration scheme is handled in \texttt{rigiddisk.c}. The functions are self-explanatory and implement routines for generating the Euler rotation matrix. At runtime, the particles in the rigid disk have their relevant quantities (acceleration, torque, etc.) rotated into the frame where the disk axis is the $z$-axis. We solve the angular part of the disk dynamics in this frame using the implicit Euler method. For the actual timestep used in the integration method, make sure you use the time calculated in \texttt{update\_halo\_positions.c}. This is the actual time elapsed as formulated\footnote{You can check that you're using the right timestep by ensuring the total time elapsed corresponds to the time elapsed in that cosmology.}, not the drift or kick steps given for the conjugate positions and velocities. We apologize for the horrible mess that this section of code has become. Perhaps it would be instructive to write your own halo tracker.

@@

As with most code developed in academia, individual researchers are largely responsible for the development, application, and maintenance of their codes. Balancing these priorities often means sacrificing on certain optimizations. Here, we list some potential improvements to the existing algorithm that might be explored.

@@

In the paper, we use a fixed timestep for the rigid disk that is based on the center-of-mass acceleration timescale. Often, this timescale is too long for the implicit Euler scheme that we use to integrate the rigid body equations. This is why we introduced a fixed \texttt{TIMESTEP\_REDUCTION\_FACTOR}. In reality, the timestep should be determined from the current magnitude of the torque and angular velocity. We would recommend experimenting with an adaptive timestep. One form that might be tried is,
\begin{equation}
\Delta t = \frac{2 \pi \alpha}{\vert \boldsymbol \omega \vert}
\end{equation}
where $\alpha$ is a constant for which lower values yield more accurate integration, and $\boldsymbol \omega$ is the angular velocity in the disk's body frame. There were a number of halos where we found the fixed timestep to be inadequate for capturing the angular evolution of the disk. This is especially an issue when the disk begins to settle in the host halo and sheds some of its angular velocity. An adaptive timestep would give us the low error required to resolve these scenarios as well as giving us efficiency when angular velocities are low.

We are also able to improve on the integration scheme. We use Implicit Euler, a first order integration scheme. By modifying the code to make use of the fact that \textsc{Gadget-3} is already using a higher order leapfrog scheme, we may be able to improve the integration accuracy.

### 2019-07-16 Thesis/Introduction.tex
Semi-final presubmission draft

\chapter{A Foray into Galactic Dynamics} \label{ch:introduction}
\newpage
%\section{Section}
%\subsection{SubSection}
%\subsubsection{SubSubSection}
%\paragraph{Paragraph}
%\subparagraph{SubParagraph}

@@

Imagine you are standing on the corner of a busy Manhattan street. Every minute that passes shows the liveliness of a dynamic system, built up over many years. On an average block, there are approximately 28,154 people in the surrounding square kilometer, and a total of 1.7 million people on the island \citep{manhattan_population_density}. In contrast, a mere 300 years ago, the population was more disperse, and numbered in the thousands \citep{history_of_nyc}. From a collection of thousands of people in a relatively rarefied population formed a massive, dynamic ecosystem that represents an epicentre of modern human innovation and industry. Manhattan is not unique; cities across the world did not appear overnight, but were constructed hierarchically. It is in a similar fashion that modern astrophysics believes our Milky Way, and all massive galaxies formed with stars and gas as their building blocks.
\begin{figure}
	\includegraphics[width=\textwidth]{../figures/2010_census}
	\caption{Population density information for the U.S. and Puerto Rico from the U.S. census bureau \citep{2010_us_census}}
\end{figure}

We believe that the Universe began with a period of rapid inflation, and that in the process of this inflation, pockets of the Universe emerged more dense than other regions. This is revealed to us through observations of the cosmic microwave background (CMB). The CMB is a portrait of the Universe when it was last opaque; before electrons, neutrons, and protons combined to form the first atoms. It shows temperature fluctations, like small population overdensities, that would eventually become cosmic cities.  While many physicists marvel at the physics that happens in these first moments of time, the story of how galaxies grow from cosmic villages contains many puzzles. This is the story of how tiny perturbations evolve to the structures astronomers see today, and it starts with the CMB.

When we look at the CMB, we are seeing the temperature distribution of matter through the radiation of baryonic matter, the matter that interacts with light. There is strong observational evidence to suggest that a large fraction of the matter present at this epoch is not interacting with light, including the power spectrum of the CMB itself \citep{transfer_fn}. The power spectrum describes the distribution of scales of the perturbations in the CMB, and presents some of the strongest evidence for the existence of dark matter, matter which does not interact with light.

Not only is there a measurable dark matter component in the Universe, it appears to be substantial. If the results from the Planck satellite are taken as truth, then dark matter comprises around 84\% of all matter in the Universe \citet{planck_2018}. Dark matter influences the evolution of galaxies not only because it is not subject to radiation-based interactions, but also because it is most of what there is. Dark matter is the glue that holds galaxies together.

If dark matter is a glue, then dark energy is the opposite. Although dark matter makes up 84\% of the matter, matter only makes up 31.5\% of the present day Universe. The remainder is composed by a small contribution from radiation, but the bulk is comprised of this unaccounted for dark energy \citep{kolb_turner,dodelson}. Dark energy, whatever it is, is likely responsible for the observation of accelerating Universe expansion from Type 1a supernova measurements \citep{reiss_1998,perlmutter_1999}. Modern cosmology is a struggle between the competing forces of dark matter and dark energy, and both have implications for the local dynamics in the Universe.

A model of the Universe one might consider is one comprised of cold dark matter, that is dark matter which is massive and slow, and one which has dark energy, described on large scales by a single parameter, $\Lambda$. This paradigm is called $\Lambda$CDM and is the prevailing model in the astronomy community. The model has been enormously successful in explaining observations, including the CMB power spectrum. While there remain open problems with $\Lambda$CDM, it appears to predict many properties on Galactic and extragalactic scales correctly \citep{kolb_turner,dodelson}. Proposals to modify $\Lambda$CDM tend to present only slight variations on the model, such as allowing for self interacting dark matter (SIDM).
\begin{figure}
	\includegraphics[width=\textwidth]{../figures/local_group}
	\caption{A map of the Local Group obtained per the license in \citet{local_group_map}.}\label{fig:local_group}
\end{figure}

Taking $\Lambda$CDM as the ground truth, the history of the Universe becomes understandable betweeen reionization and today. Dark matter, which does not interact through any force but gravity, retains a perturbed structure after inflation. The density perturbations collapse, forming potential wells of dark matter. The baryons, interacting through baryon acoustic oscillations (BOA), are delayed in falling into the potential wells made by the dark matter. As the baryons fall into the centers of the potential wells over time, the cores of the dark matter distributions, comprised of spheroidal units called halos, contract. After infall into a dark matter halo, the baryons cool to the point where stars are able to form. This is how the first galaxies formed.

@@

\section{Simulations as a Tool For Understanding Problems in $\Lambda$CDM}

Many of the hardest questions in Galactic astrophysics stem from the fact that dark matter is not directly observable. As an example, what is the underlying smooth distribution of dark mass? How many dark matter subhalos should the Milky Way have? What is the effect on large scales if we choose something other than $\Lambda$CDM? While these questions are not easily answerable by direct observation, we can complement indirect observations with a solid theoretical understanding to reach some confident answers.

Simulations are the complementary theoretical understanding of galactic and extragalactic systems. When a simulation is performed, theoretical input is considered. This includes your model of cosmology, or your model of the galaxy, the nature of dark matter, etc. All of this knowledge is reduced to a fluid system which may be solved through numerical techniques on a computer.

Over the last four decades, many tools have been developed to study the structure of the present day Universe. These tools have become increasingly sophisticated as computation power and algorithms research advance. One tool is an \textit{ab initio} cosmological simulation, which simulates the entire formation history of the Universe from some informed initial density field.  This initial density field is derived from the primordial power spectrum, and a sample can be drawn and evolved with periodic boundary conditions \citep{music}. If done correctly, this creates a representative sample of many structures in a $\Lambda$CDM Universe. Halos evolving in these simulations feel the full effect of the completely simulated cosmological environment. Such simulations stand in contrast to isolated galaxy simulations, which seek to explain the detailed dynamics arising from generic effects in individual galaxies. These simulations start with a prescription for what an equilibrium galaxy looks like, and study the evolution of galaxies initialized from these prescriptions. While missing some critical cosmological physics, these simulations may be run at very high resolution.

@@

The Core-Cusp Problem and Missing Satellite Problem have received a lot of attention in the literature because of the potential problems they pose for theoretical cosmology. It is clear that to accurately recover the global properties of substructure on large scales, dark matter must be cold. Warm and hot dark matter models\footnote{Hot dark matter is comprised of matter moving at or near the speed of light (e.g. neutrinos), and warm dark matter is somewhere in between cold and hot dark matter.} simply do not yield substructure populations consistent with observations. SIDM and fuzzy models, where the dark matter can interact on small scales, have been proposed as slight modifications to the $\Lambda$CDM paradigm, with some success. Despite the presence of alternate theories of cosmology, a large concerted effort has been made to understand discrepancies as a byproduct of baryonic physics.  Adequeate explanation of halo and substructure response to stellar and gas content would mean that alternate cosmologies are not necessary to explain the Core-Cusp Problem and Missing Satellite Problem.

Initially, simulations included only dark matter, but have evolved over the last three decades to consider very sophisticated subgrid models of gas physics and star formation \citep{LucySPH,CenOstriker92,KatzSPH96,SpringelMultiphase03,Stinson2006}.  These simulations can be used to realize the galaxy formation predicted by current theories. State of the art unigrid (single resolution) cosmological simulations struggle to compete with the resolution attainable by isolated galaxy simulations \citep{Illustris,Eagle}, but they can still be used to make broad statements about consistency with $\Lambda$CDM. Broadly speaking, it is not clear at this stage whether or not apparent issues with $\Lambda$CDM can be resolved through these models, but they present a promising approach as computing power is sure to increase.

@@

Up until this point, we have focused on the halo. This is the most massive component of our galaxy, holding a complicated geometry. The nature of this geometry is not directly probable, and we must rely on indirect methods to infer information about the halo. In addition to the dark matter halo, our Galaxy is composed primarily of a central spheroidal bulge, a thin disk, and a more diffuse thick disk. There is also a stellar component of the halo, which contains remnants of old mergers and star clusters known as globular clusters. This section is organized by talking about tidal disruption of subhalos in the Galactic halo, the dynamics of the central part of the Galaxy, and vertical structure in the Galactic disk. How the cosmological environment effects these observables is the bulk of this thesis, and we motivate here why these aspects are the subject of focus.

@@

More streams have been discovered since then, and a few notable ones are the Orphan Stream \citep{grillmair_2006,belokurov_2007,newberg_2010}, remarkable for its apparent lack of progenitor, TriAnd, A13, and the Monoceros Ring. This is by no means a complete accounting of known stellar streams. Something on the order of more than a dozen streams are known to exist in the Milky Way alone \citep{sanders_binney_2013_a}. While streams alone have been useful in getting some broad contraints on the Galactic potential and geometry of the halo, complications arise from the fact that streams do not follow orbits exactly \citet{sanders_binney_2013_a} and the fact that the Milky Way potential is predicted from theory to be time dependent.

@@

\section{Bridging the Gap}

@@

\citet{bauer2018a} improved on the scheme even further by increasing the realism of the growth phase by modeling the disc as a rigid body. The results they found were qualitatively similar to \citet{debuhr_2012} and \citet{ys_2015}  in terms of disc dynamics. The live phase was also improved by using the \textsc{GalactICS} code which actually samples a three-integral DF. It is using this tool that this thesis is constructed upon.

@@

The thesis is organized in the following manner. In Chapter~\ref{ch:background}, we reduce the necessary theoretical background to understand this work to a succinct summary. Following this, in Chapter~\ref{ch:paper_i}, we explore the first peer-reviewed paper on which this thesis is based. It outlines the disk insertion scheme and comments on how the dynamic nature of the Milky Way's stellar disk should affect its inner halo. Chapter~\ref{ch:paper_ii} presents the second published paper which focuses on the question of how stable disks are against forming bars in a cosmological setting. The results discussed in Chapter~\ref{ch:paper_ii} are contextualized with the results of isolated galaxy simulations.

>>>>

\chapter{Introduction} \label{ch:introduction}
\newpage
%\section{Section}
%\subsection{SubSection}
%\subsubsection{SubSubSection}
%\paragraph{Paragraph}
%\subparagraph{SubParagraph}

@@

Imagine you are standing on the corner of a busy Manhattan street. Every minute that passes shows the liveliness of a dynamic system, built up over many years. On an average block, there are approximately 28,154 people in the surrounding square kilometer, and a total of 1.7 million people on the island \citep{manhattan_population_density}. In contrast, a mere 300 years ago, the population was more disperse, and numbered in the thousands \citep{history_of_nyc}. Collections of thousands of people, initially spread out, came together to form a massive, dynamic ecosystem that represents an epicentre of modern human innovation and industry. Manhattan is not unique; cities across the world did not appear overnight, but were constructed hierarchically. It is in a similar fashion that modern astrophysics believes our Milky Way, and all massive galaxies formed with stars and gas as their building blocks.

We believe that the Universe began with a period of rapid inflation, and that in the process of this inflation, pockets of the Universe emerged more dense than other regions. This is revealed to us through observations of the cosmic microwave background (CMB). The CMB is a portrait of the Universe when it was last opaque; before electrons, neutrons, and protons combined to form the first atoms. It shows temperature fluctuations, like small population overdensities, that would eventually become cosmic cities.  While many physicists marvel at the physics that happens in these first moments of time, there is another story of how cosmic villages come together to form cities. This is the story of how tiny perturbations evolve to the structures astronomers see today.

When we look at the CMB, we are seeing the temperature distribution of matter through the radiation of baryonic matter, the matter that interacts with light. There is strong observational evidence to suggest that a large fraction of the matter present at this epoch is not interacting with light. This evidence is found in the power spectrum of the CMB, a function that describes the distribution of scales of the perturbations in the CMB \citep{transfer_fn}.

This dark matter appears to be present in the current epoch. Evidence of dark matter in modern-day galaxy clusters was first observed by \citet{zwicky_1933}, who inferred that the mass of the Coma Cluster needed to be approximately ten times higher than observed to explain the motions of the constituent galaxies. A convincing case for dark matter in individual galaxies was presented by \citet{rubin_1980}, who show that the rotational speeds of stars in the disks of galaxies were too high to be explained by observed mass.

Based on enormous observational evidence, the dark matter component of the Universe appears to be substantial. If the results from the Planck satellite are taken as truth, then dark matter comprises around 84\% of all matter in the Universe \citep{planck_2018}. Dark matter influences the evolution of galaxies not only because it is not subject to radiation-based interactions, but also because it is most of what there is. Dark matter is the glue that holds galaxies together.

Although dark matter makes up 84\% of the matter, matter only makes up 31.5\% of the present day Universe's total energy density. The remainder contains a modest contribution from radiation, but the bulk is comprised of unaccounted dark energy \citep{kolb_turner,dodelson, BT}. Dark energy, whatever it is, is responsible for the accelerating expansion of the Universe, something we first learned from from observations of Type 1a supernova \citep{reiss_1998,perlmutter_1999}. Modern cosmology is a struggle between the competing forces of dark matter and dark energy, and both have implications for the local dynamics in the Universe.

A model of the Universe one might consider is one comprised of cold dark matter, that is dark matter which is massive and slow, and one which has dark energy, described on large scales by a single parameter, $\Lambda$. This paradigm is called $\Lambda$CDM and is the prevailing model in the astronomy community. The model has been enormously successful in explaining observations, including the CMB power spectrum. While there remain open problems with $\Lambda$CDM, it appears to predict many properties on Galactic and extragalactic scales correctly \citep{kolb_turner,dodelson, BT}. Proposals to modify $\Lambda$CDM tend to present only slight variations on the model, such as allowing for self interacting dark matter (SIDM).
\begin{figure}
	\includegraphics[width=\textwidth]{../figures/local_group}
	\caption{A map of the Local Group obtained per the license in \citet{local_group_map}.}\label{fig:local_group}
\end{figure}

\subsection{The Story of the Milky Way}
Taking $\Lambda$CDM as the ground truth, we can begin to understand how galaxies evolved from tiny villages. Dark matter, which does not interact through any force but gravity, retains a perturbed structure after inflation. The density perturbations collapse, forming potential wells of dark matter \citep{taylor_2011}. The baryons, interacting through baryon acoustic oscillations (BOA), are delayed in falling into the potential wells made by the dark matter. As the baryons fall into the centers of the potential wells over time, the cores of the dark matter distributions, comprised of spheroidal units called halos, contract. After infall into a dark matter halo, the baryons cool to the point where stars are able to form. This is how the first galaxies formed.

@@

\section{Understanding $\Lambda$CDM with Galaxy Simulations}

Many of the hardest questions in Galactic astrophysics stem from the fact that dark matter is not directly observable. As an example, what is the underlying smooth distribution of dark mass? How many dark matter subhalos should the Milky Way have? What is the effect on large scales if we choose something other than $\Lambda$CDM? While these questions are not easily answerable by direct observation, we can complement indirect observations with a solid theoretical understanding to reach confident answers.

Simulations are the complementary theoretical understanding of galactic and extragalactic systems. When a simulation is performed, theoretical input is considered. This includes a model of cosmology, or a model of the galaxy, the nature of dark matter, etc. All of this knowledge is reduced to a fluid system which may be solved through numerical techniques on a computer.

Over the last four decades, many tools have been developed to study the structure of the present day Universe. These tools have become increasingly sophisticated as computation power and algorithms research advance. One tool is an \textit{ab initio} cosmological simulation, which simulates the entire formation history of the Universe from some informed initial density field.  This initial density field is derived from the primordial power spectrum and early-Universe baryon physics, and a sample can be drawn and evolved with periodic boundary conditions \citep{music}. If done correctly, this creates a representative sample of many structures in a $\Lambda$CDM Universe. Halos evolving in these simulations feel the full effect of the completely simulated cosmological environment. Such simulations stand in contrast to isolated galaxy simulations, which seek to explain the detailed dynamics arising from generic effects in individual galaxies. These simulations start with a prescription for what an equilibrium galaxy looks like, and study the evolution of galaxies initialized from these prescriptions. While missing some critical cosmological physics, these simulations may be run at very high resolution.

@@

The Core-Cusp Problem and Missing Satellite Problem have received a lot of attention in the literature because of the potential problems they pose for theoretical cosmology. It is clear that to accurately recover the global properties of substructure on large scales, dark matter must be cold. Warm and hot dark matter models\footnote{Hot dark matter is comprised of matter moving at or near the speed of light (e.g. neutrinos), and warm dark matter is somewhere in between cold and hot dark matter.} simply do not yield substructure populations consistent with observations \citep[for example]{alternate_dm}. SIDM and fuzzy models, where the dark matter can interact on small scales, have been proposed as slight modifications to the $\Lambda$CDM paradigm, with some success \citep[for example]{hui_2017}.

Despite the presence of alternate theories of cosmology, a large concerted effort has been made to understand discrepancies as a byproduct of baryonic physics.  Adequate explanation of halo and substructure response to stellar and gas content would mean that alternate cosmologies are not necessary to explain the Core-Cusp Problem and Missing Satellite Problem. Initially, \textit{ab initio} simulations included only dark matter, but have evolved over the last three decades to consider very sophisticated subgrid models of gas physics and star formation \citep{LucySPH,CenOstriker92,KatzSPH96,SpringelMultiphase03,Stinson2006}.  These simulations can be used to realize the galaxy formation predicted by current theories. State of the art unigrid (single resolution) cosmological simulations struggle to compete with the resolution attainable by isolated galaxy simulations \citep{Illustris,Eagle}, but they can still be used to make broad statements about consistency with $\Lambda$CDM. Broadly speaking, it is not clear at this stage whether or not apparent issues with $\Lambda$CDM can be resolved through these models, but they present a promising approach as computing power is sure to increase.

@@

Since dark matter is not directly probable, we must use indirect methods to infer information about the halo. In addition to the dark matter halo, our Galaxy is composed of baryons in the form of a central spheroidal bulge, a thin gas disk, a thin stellar disk, and a more diffuse thick stellar disk. There is also a stellar component of the halo, which contains remnants of old mergers and star clusters known as globular clusters. This section is organized by talking about tidal disruption of subhalos in the Galactic halo, the dynamics of the central part of the Galaxy, and vertical structure in the Galactic disk. How the cosmological environment effects these observables is the bulk of this thesis, and we motivate here why these aspects are the subject of focus.

@@

More streams have been discovered since then, and a few notable ones are the Orphan Stream \citep{grillmair_2006,belokurov_2007,newberg_2010}, remarkable for its apparent lack of progenitor, TriAnd, A13, and the Monoceros Ring. This is by no means a complete accounting of known stellar streams. Something on the order of more than a dozen streams are known to exist in the Milky Way alone \citep{sanders_binney_2013_a}. While streams alone have been useful in getting some broad constraints on the Galactic potential and geometry of the halo, complications arise from the fact that streams do not follow orbits exactly \citet{sanders_binney_2013_a} and the fact that the Milky Way potential is predicted from theory to be time dependent.

@@

\section{Disk Insertion Techniques for Cosmology}

@@

\citet{bauer2018a} improved on the scheme even further by increasing the realism of the growth phase by modelling the disc as a rigid body. The results they found were qualitatively similar to \citet{debuhr_2012} and \citet{ys_2015}  in terms of disc dynamics. The live phase was also improved by using the \textsc{GalactICS} code which actually samples a three-integral DF. It is using this tool that this thesis is constructed upon.

@@

The thesis is organized in the following manner. In Chapter~\ref{ch:background}, we reduce the necessary theoretical background to understand this work to a succinct summary. Following this, in Chapter~\ref{ch:paper_i}, we explore the first peer-reviewed paper on which this thesis is based. It outlines the disk insertion scheme and comments on how the dynamic nature of the Milky Way's stellar disk should affect its inner halo. Chapter~\ref{ch:implementation} is intended to clarify some implementation details of the algorithm in Chapter~\ref{ch:paper_i}. Chapter~\ref{ch:paper_ii} presents the second published paper which focuses on the question of how stable disks are against forming bars in a cosmological setting. A final paper is presented in Chapter~\ref{ch:paper_iii}, where we look at how disks form vertical structure in a cosmological setting. We conclude with summaries and ideas for future work in Chapter~\ref{ch:conclusion}.

### 2019-07-16 Thesis/Paper_I.tex
Writing for May 9
Semi-final presubmission draft

\chapter{A Method for Studying Discs in Cosmological Haloes}

\section{Abstract}

@@

%\def \last_written_page{\thepage}
\bibliographystyle{apalike}
\bibliography{bibliography_paper_i.bib}
%\last_page_of_biblio
%\def \last_written_page{\thepage}

>>>>

\chapter{A Method for Studying Discs in Cosmological Haloes} \label{ch:paper_i}
\textbf{This chapter contains a reproduction of a paper published in  Monthly Notices of the
Royal Astronomical Society as: Jacob S. Bauer, Lawrence M. Widrow, and Denis Erkal. Disc-halo interactions in $\Lambda$CDM. \textit{Monthly Notices of the Royal Astronomical Society}, 476:198-209,
2018.}
\newpage
\section{Abstract}

@@

\section*{Appendix A: Euler's Equations in Comoving Coordinates} \label{sec:derivation}

The time-evolution of the angular momentum vector ${\bf L}$ of a rigid
body acted upon by a torque $\boldsymbol{\tau}$ is given by

\begin{equation}
  \left(\frac{{d} \bf{L}}{{d} t}\right)_f = \left(\frac{{d} \bf{L}}{{d} t}\right)_b  + \boldsymbol \omega \times \bf{L}
  = \boldsymbol{\tau}
\end{equation}

\noindent where the subscripts $f$ and $b$ denote the frame of the
simulation box and the body frame, respectively.  In physical
coordinates, ${\bf L} = {\bf r \times p}$.  Alternatively, we can
write ${\bf L} = {\bf s\times q}$ where ${\bf s} = a^{-1}{\bf r}$
refer to comoving coordinates and ${\bf q} = a^2\dot{\bf s}$ is the
conjugate momentum to ${\bf s}$ (see \citet{QuinnIntegrators}).

For a rigid body, the components of the angular momentum are given by $L_i = I_{ij} \omega_j$ where $i,j$ run over $x,\,y,\,z$ and there is an implied sum over $j$.  Since \textsc{gadget-3} uses comoving coordinates, we write $I_{ij} = a^2 J_{ij}$ where $J$ is the moment of inertia tensor written in terms of the comoving coordinates, ${\bf s}$, rather than the physical coordinates, ${\bf r}$.  For convenience, we define a ``comoving'' angular velocity $\boldsymbol{\varpi} = a^{-2}\boldsymbol{\omega}$.  We then have $L_i = J_{ij} \varpi_j$.  Note that because of the symmetry of our disc, the moment of inertia tensor is diagonal with $J_{xx} = J_{yy}= J_{zz} \equiv J/2$.  The equations of motion for the Euler angles and the disc angular velocity are then given by the standard Euler equations of rigid body dynamics, modified to account for the time-dependence of the disc's moment of inertia:

\begin{equation}
\frac{d\phi}{dt} = a^{-2}\sin^{-1}{\theta} \left (\varpi_x\sin(\psi) + \varpi_y \cos(\psi)\right )~,
\end{equation}
\begin{equation}
\frac{d\theta}{dt} = a^{-2}\left (  \varpi_1\cos(\psi) - \varpi_y \sin(\psi)\right )~,
\end{equation}
\begin{equation}
J\frac{\varpi_x}{dt} + \varpi_x\frac{dJ}{dt}
+ J\varpi_y\varpi_z
=  \tau_x~,
\end{equation}
and
\begin{equation}
\frac{\varpi_y}{dt} + \varpi_y\frac{dJ}{dt}
- J\varpi_x\varpi_z
=  \tau_y~.
\end{equation}

\noindent We have omitted the equations for $\psi$ (rotations in the body frame about the symmetry axis) and $\varpi_z$ since these are determined directly from Eq.\,\ref{eq:omega3}.   
%\def \last_written_page{\thepage}
\bibliographystyle{apalike}
\bibliography{bibliography_paper_i.bib}
%\last_page_of_biblio
%\def \last_written_page{\thepage}
